---
title: "1000plus collateral flow & perfusion project (Kersten Villringer)"
author: "Ahmed Khalil"
date: "21 May 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# necessary packages
library(robustbase)
library(knitr)
library(xtable)
library(Hmisc)

# load data
allpts <- read.table("Carmela_Project.csv", header = T, sep=",")

# create necessary variables (reperfusion / infarct growth binarized cut-offs)
allpts$reperf[allpts$relTmax_change<=-50] <- "Yes"
allpts$reperf[allpts$relTmax_change>-50] <- "No"
allpts$reperf <- factor(allpts$reperf)

# recode recanalization variable
allpts$Rekan[allpts$Rekan==0] <- "None" # no recanalization
allpts$Rekan[allpts$Rekan==1] <- "Full" # full recanalization
allpts$Rekan[allpts$Rekan==2] <- "Partial" # partial recanalization
allpts$Rekan <- factor(allpts$Rekan)

# recode other categorical variables
allpts$sex[allpts$sex==1] <- "Male"
allpts$sex[allpts$sex==2] <- "Female"
allpts$sex <- factor(allpts$sex)

allpts$iv_tPR[allpts$iv_tPR==0] <- "No"
allpts$iv_tPR[allpts$iv_tPR==1] <- "Yes"
allpts$iv_tPR <- factor(allpts$iv_tPR)

allpts$circulation[allpts$circulation==1] <- "Anterior"
allpts$circulation[allpts$circulation==2] <- "Posterior"
allpts$circulation <- factor(allpts$circulation)

# remove certain variables
allpts$no_recan_1 <- c()
allpts$VAR00002 <- c()
allpts$TSI_dich <- c()
```

### Descriptive statistics
```{r, echo= FALSE, fig.width=6, fig.height=6, fig.align='center',dpi=300, warning=F}
# descriptive stats
describe(allpts[,-1])
```


### (1)How many patients with reperfusion are recanalizers or non-recanalizers and (2.) is there a difference in reperfusion status between the groups?
```{r, echo= FALSE, fig.width=6, fig.height=6, fig.align='center',dpi=300, warning=F}


# cross-tabulation of recanalization/reperfusion
kable(table(allpts$reperf, allpts$Rekan))
barplot(table(allpts$reperf, allpts$Rekan), legend.text = c("No reperfusion", "Reperfusion"), col=c("#1b9e77","#d95f02"), xlab="Recanalization", ylab="Number of patients", xpd=F,args.legend = list(x = "topleft"), ylim=c(0,80))

# test for differences between recanalization and reperfusion

```


### Performing ANCOVA (! Assumptions)
```{r, echo= FALSE, fig.width=6, fig.height=6, fig.align='center',dpi=300, warning=F}
# check assumptions of ANCOVA
par(mfrow=c(2,2))
plot(lm(allpts$relTmax_change~allpts$Rekan + allpts$TSI *allpts$iv_tPR))

# perform ANCOVA
summary(lm(allpts$relTmax_change~allpts$Rekan + allpts$TSI + allpts$iv_tPR))
```

### (3.) Is there a difference in perfusion changes between patients with anterior and posterior circulation strokes?
```{r, echo= FALSE, fig.width=6, fig.height=6, fig.align='center',dpi=300, warning=F}

# test if circulation has influence on reperfusion in non-recanalizers
summary(lmrob(allpts[allpts$Rekan=="None",]$relTmax_change~allpts[allpts$Rekan=="None",]$circulation))

# test if circulation has influence on reperfusion in non-recanalizers ADNJUSTING FOR iv TPA
summary(lmrob(allpts[allpts$Rekan=="None",]$relTmax_change~allpts[allpts$Rekan=="None",]$circulation*allpts[allpts$Rekan=="None",]$iv_tPR))

# test if circulation has influence on reperfusion in recanalizers (full OR partial)
summary(lmrob(allpts[allpts$Rekan=="Full"|allpts$Rekan=="Partial",]$relTmax_change~allpts[allpts$Rekan=="Full"|allpts$Rekan=="Partial",]$circulation))
```

### (4.) Is there an influence of rTPA, TOO, recanalization on perfusion changes?
```{r, echo= FALSE, fig.width=6, fig.height=9, fig.align='center',dpi=300, warning=F}
# check assumptions of robust GLM
par(mfrow=c(3,2))
plot(lmrob(allpts$relTmax_change~allpts$Rekan + allpts$circulation + allpts$TSI + allpts$iv_tPR))

# perform robust GLM
summary(lmrob(allpts$relTmax_change~allpts$Rekan + allpts$circulation + allpts$TSI + allpts$iv_tPR))
par(mfrow=c(1,1))
boxplot(relTmax_change~Rekan, data=allpts, col=c("#1b9e77","#1b9e77"), ylab="Relative Tmax change")

```

### (5.) Is there a difference in clinical outcome (mRS day90) in reperfused vs non-reperfused recanalizers and non-recanalizers?
```{r, echo= FALSE, fig.width=8, fig.height=6, fig.align='center',dpi=300, warning=F}
# perform robust GLM
summary(lmrob(allpts$mRS_d90~allpts$Rekan*allpts$reperf + allpts$NIHSS_ad + allpts$iv_tPR +allpts$TSI + allpts$rel_DWIgrowth))
boxplot(mRS_d90~Rekan + reperf, data=allpts, col=c("#1b9e77","#1b9e77","#1b9e77","#d95f02","#d95f02","#d95f02"), ylab="mRS day 90")
legend("topleft", legend = c("Non-reperfused","Reperfused"), fill=c("#1b9e77","#d95f02"))
```

### (6.) Is there a difference in imaging outcome (infarct growth) in reperfused vs non-reperfused recanalizers and non-recanalizers?
```{r, echo= FALSE, fig.width=8, fig.height=6, fig.align='center',dpi=300, warning=F}
# perform robust GLM
summary(lmrob(allpts$rel_DWIgrowth~allpts$Rekan*allpts$reperf + allpts$NIHSS_ad + allpts$iv_tPR +allpts$TSI))
boxplot(rel_DWIgrowth~Rekan + reperf, data=allpts,col=c("#1b9e77","#1b9e77","#1b9e77","#d95f02","#d95f02","#d95f02"), ylab="Relative DWI growth", ylim=c(-100,300))
legend("topleft", legend = c("Non-reperfused","Reperfused"), fill=c("#1b9e77","#d95f02"))

summary(lmrob(allpts$DWIgrowth~allpts$Rekan*allpts$reperf + allpts$NIHSS_ad + allpts$iv_tPR +allpts$TSI))
boxplot(DWIgrowth~Rekan + reperf, data=allpts,col=c("#1b9e77","#1b9e77","#1b9e77","#d95f02","#d95f02","#d95f02"), ylab="Absolute DWI growth", ylim=c(0,100))
legend("topleft", legend = c("Non-reperfused","Reperfused"), fill=c("#1b9e77","#d95f02"))

```