---
title: "1000plus collateral flow & perfusion project"
author: "Ahmed Khalil"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: 
    toc: true
    toc_float: true
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# necessary packages
library(robustbase)
library(knitr)
library(xtable)
library(Hmisc)
library(TOSTER)
library(stargazer)
library(summarytools)
library(magrittr)
library(kableExtra)
library(effects)
library(pwr)
library(papeR)
library(robust)
library(reshape2)

# DEFINE WHETHER YOU WANT EFFECTS PLOTS FOR MIXED MODELS
effplot = "Y"

# Define whether (T) or not (F) to show outliers in boxplots
OL = F

#if (file.exists("allpts.Rdata")){
  #("allpts.Rdata")
#} else {
# load data
allpts <- read.table("all_data_v3.csv", header = T, sep=";")
colnames(allpts)[1] <- "ID"

# load volumetric data
auto_pts <- read.table("ALLVOLS_automated_AK_20180605.csv", header=T, sep=";") # patients automatically delineated
manual_pts <- read.table("ALLVOLS_manual_AK_20180605.csv", header=T, sep=";") # patients manually delineated

# remove # of voxels, keep only volumes (in mm3)
auto_pts <- auto_pts[,c(1,3,5,7,9,11,13)]
manual_pts <- manual_pts[,c(1,3,5,7,9,11,13)]

# convert volumes to mL (cm3)
auto_pts[,-1] <- auto_pts[,-1]*0.001
manual_pts[,-1] <- manual_pts[,-1]*0.001

# combine manual and automated volumes
vols_all_pts <- rbind(auto_pts, manual_pts)

# exclude subject 86 (not in Kersten's main dataset)
vols_all_pts <- vols_all_pts[!vols_all_pts$ID=="86",]

# calculate HIR for days 1 and 2 
vols_all_pts$HIR1 <- vols_all_pts$Tmax_D1_.8_volume/vols_all_pts$Tmax_D1_.2_volume
vols_all_pts$HIR2 <- vols_all_pts$Tmax_D2_.8_volume/vols_all_pts$Tmax_D2_.2_volume

# merge volumes and HIR with the rest of the data
allpts <- merge(allpts, vols_all_pts, by = "ID")

# rename Tmax volume variables to include threshold (>6s)
allpts$Tmax_D1_.6_volume <- allpts$Tmax_vol_d1
allpts$Tmax_D2_.6_volume <- allpts$Tmax_vol_d2

# create reperfusion variables
allpts$Tmax_6_reperf_abs <- allpts$Tmax_D1_.6_volume - allpts$Tmax_D2_.6_volume
allpts$Tmax_6_reperf_rel <- (allpts$Tmax_6_reperf_abs/allpts$Tmax_D1_.6_volume)*100
# replace infinity values (divide by zero where D1 Tmax volume = 0) with NA
is.na(allpts$Tmax_6_reperf_rel)<-sapply(allpts$Tmax_6_reperf_rel, is.infinite)

# recode recanalization variable
allpts$Rekan[allpts$Rekan==0] <- "None" # no recanalization
allpts$Rekan[allpts$Rekan==1] <- "Full" # full recanalization
allpts$Rekan[allpts$Rekan==2] <- "Partial" # partial recanalization
allpts$Rekan <- factor(allpts$Rekan)
allpts$Rekan <- factor(allpts$Rekan,levels(allpts$Rekan)[c(2,3,1)])

# recode other categorical variables
allpts$sex[allpts$sex==1] <- "Male"
allpts$sex[allpts$sex==2] <- "Female"
allpts$sex <- factor(allpts$sex)

allpts$iv_tPR[allpts$iv_tPR==0] <- "No"
allpts$iv_tPR[allpts$iv_tPR==1] <- "Yes"
allpts$iv_tPR[allpts$iv_tPR==2] <- "Yes" 

allpts$iv_tPR <- factor(allpts$iv_tPR)

allpts$circulation[allpts$circulation==1] <- "Anterior"
allpts$circulation[allpts$circulation==2] <- "Posterior"
allpts$circulation <- factor(allpts$circulation)

allpts$higashida_change <- allpts$higashida_D2-allpts$higashida_D1
allpts$higashida_D1 <- factor(allpts$higashida_D1)
allpts$higashida_D2 <- factor(allpts$higashida_D2)
allpts$higashida_change <- factor(allpts$higashida_change)

# remove certain variables
allpts$no_recan_1 <- c()
allpts$VAR00002 <- c()
allpts$TSI_dich <- c()

# merge with variables defining occlusion location
occ_classfn <- read.table("occlusion_classfn.csv", header = T, sep=";")
allpts <- merge(allpts, occ_classfn, by = "ID")

# correctly code NaNs
allpts[allpts=="NaN"]<-NA

# save data 
save(allpts, file = "allpts.Rdata")

#}


```
### Methods
**Reperfusion** is defined as the relative Tmax >6s change (% of baseline).  
Surrogates for good collateral flow in this study are:  
  * Degree of reperfusion in non-recanalizers  
  * The hypoperfusion intensity ration (HIR) at baseline (lower is better), defined according to:  
*Bang, O. Y. et al. (2008) ‘Determinants of the distribution and severity of hypoperfusion in patients with ischemic stroke.’, Neurology*  
  * The modified ASITN/SIR Collateral Flow Grading System (adapted for DSC-MRI, aka "Higashida score"), according to:  
*Campbell BC V, Christensen S, Tress BM, Churilov L, Desmond PM, Parsons MW, Alan Barber P, Levi CR, Bladin C, Donnan GA, Davis SM (2013) Failure of collateral blood flow is associated with infarct growth in ischemic stroke. J Cereb Blood Flow Metab* and dichotomized into good (3-4) and poor (0-2) according to: 
*Galinovic I, Kochova E, Khalil A, Villringer K, Piper SK, Fiebach JB (2018) The ratio between cerebral blood flow and Tmax predicts the quality of collaterals in acute ischemic stroke. PLoS One 13:e0190811*

The outcome measures in this study are:  
* Extent of reperfusion - relative Tmax >6s change (%)  
* Modified Rankin Scale at Day 90  
* Absolute infarct growth on DWI between days 1 and 2 (mL)  
* Absolute infarct growth between days 1 (DWI) and 5 (FLAIR) (mL)

All regression analyses are performed using a robust GLM ("High Breakdown And High Efficiency Robust Linear Regression") using the R function "lmrob" from the "robustbase" package.  
Citation of package:  
*Martin Maechler, Peter Rousseeuw, Christophe Croux, Valentin Todorov, Andreas Ruckstuhl, Matias Salibian-Barrera, Tobias Verbeke, Manuel Koller, Eduardo L. T. Conceicao  and Maria Anna di Palma (2018). robustbase: Basic Robust Statistics R package version 0.93-0. URL http://CRAN.R-project.org/package=robustbase *   
Citation of robust GLM:  
*Yohai, V. J. (1987). High breakdown-point and high efficiency robust estimates for regression. The Annals of Statistics, 642-656.*

For visualization (not analysis) purposes, outliers are **not** shown on the boxplots. 

**Power analysis:**
`r pwrtest <- pwr.f2.test(u= 8, f2=0.1, sig.level = 0.05, power=0.8)`

Sample size needed to detect a minimum effect size (R2/1-R2)  of `r pwrtest$f2` at significance level of `r pwrtest$sig.level` and power of `r pwrtest$power` with `r pwrtest$u` predictors (including interactions) = `r round(pwrtest$v + pwrtest$u + 1)`. Performed with the R package "pwr".  
Citation of package:  
*Stephane Champely (2018). pwr: Basic Functions for Power Analysis. R package version 1.2-2. https://CRAN.R-project.org/package=pwr *
  
### Summary of main results  
* Higashida score and recanalization are independent predictors of reperfusion. See [Predictors of extent of reperfusion]
* Collateral flow (HIR or Higashida score) does not predict reperfusion in full recanalizers. See [Predictors of extent of reperfusion in full recanalizers]
* Collateral flow (HIR or Higashida score) does not predict reperfusion in non-recanalizers. See [Predictors of extent of reperfusion in non-recanalizers]
* Higashida score independently predicts reperfusion in partial recanalizers. See [Predictors of extent of reperfusion in partial recanalizers]
* NIHSS on admission and recanalization independently predict 90 day mRS. Collateral flow and reperfusion do not. See [Predictors of clinical outcome]
* Collateral flow (both Higashida and HIR), NIHSS on admission, and reperfusion in partial recanalizers independently predict early infarct growth (D1 - D2). See [Predictors of absolute infarct growth (D1 - D2)]
* Collateral flow (both Higashida and HIR), NIHSS on admission, and reperfusion in partial recanalizers independently predict late infarct growth (D1 - D5). See [Predictors of absolute infarct growth (D1 - D5)]


### Descriptive statistics
```{r, echo= FALSE, fig.width=6, fig.height10, fig.align='center',dpi=300, warning=F}
# descriptive stats
kable(summarize(allpts, type="numeric"))

summarize(allpts, type="factor", variables=c("higashida_D1", "higashida_D2", "Rekan", "circulation", "sex", "iv_tPR"))

# visualize changes in Higashida over time
# create Higashida data frame
higa_mat <- cbind(allpts$higashida_D1, allpts$higashida_D2)
# sort Higashida data frame based on D1
higa_mat_sort <- higa_mat[order(higa_mat[,1]),]
# make Higashida data frame long
higa_long <- melt(higa_mat_sort)
# give meaningful names to columns
colnames(higa_long) <- c("Subject","Day","Higashida")
higa_long$Higashida <- factor(higa_long$Higashida)
# plot as heat map (i.e. Lasagne plot)
ggplot(higa_long, aes(x=Day, y=Subject, fill=Higashida)) + geom_tile(colour="black") + scale_x_discrete(limits=c(1,2)) + scale_fill_manual(values=c("#d7191c", "#fdae61", "#ffffbf", "#abd9e9", "#2c7bb6", "#000000"),name="Higashida score", labels = c("0","1","2","3","4","NA")) 
                    
```
*Plot showing the change in Higashida scores over time for each patient.*
<style>

table, td, th {
  border: none;
  padding-left: 1em;
  padding-right: 1em;
  min-width: 50%;
  margin-left: auto;
  margin-right: auto;
  margin-top: 1em;
  margin-bottom: 1em;
}

</style>

### Predictors of extent of reperfusion 
```{r, echo= FALSE, fig.width=3, fig.height=3, fig.align='center',dpi=300, warning=F, results="asis"}
model_A <- lmrob(Tmax_6_reperf_rel~NIHSS_ad + iv_tPR + circulation, data=allpts, setting = "KS2014")
model_B <- lmrob(Tmax_6_reperf_rel~NIHSS_ad + iv_tPR + Rekan + circulation, data=allpts, setting = "KS2014")
model_C <- lmrob(Tmax_6_reperf_rel~NIHSS_ad + iv_tPR + Rekan*HIR1 + circulation, data=allpts, setting = "KS2014")
model_D <- lmrob(Tmax_6_reperf_rel~NIHSS_ad + iv_tPR + Rekan*higashida_D1_bin + circulation, data=allpts, setting = "KS2014")

pdf("reperf_RekanHIR.pdf", width = 5, height = 5)
plot(effect(mod = model_C, term = "Rekan*HIR1"), multiline=T, ci.style="bars")
dev.off()
pdf("reperf_RekanHig.pdf", width = 5, height = 5)
plot(effect(mod = model_D, term = "Rekan*higashida_D1_bin"), multiline=T, ci.style="bars")
dev.off()

stargazer(model_A, model_B, model_C, model_D, type="html", align = F, style="all", single.row=T)
stargazer(anova(model_A,model_B), type="html", align = F, style="all", single.row=T, title = "1 vs 2")
stargazer(anova(model_A,model_C), type="html", align = F, style="all", single.row=T, title = "1 vs 3")
stargazer(anova(model_A,model_D), type="html", align = F, style="all", single.row=T, title = "1 vs 4")

stargazer(model_C, model_D, type="html", align = F, style="all", single.row=T, covariate.labels = c("NIHSS on admission", "IV tPA (yes)", "Recanalization (partial)","Recanalization (full)", "HIR", "Higashida (poor)", "Circulation (posterior)", "Recanalization (partial)*HIR", "Recanalization (full)*HIR", "Recanalization (partial)*Higashida (poor)", "Recanalization (full)*Higashida (poor)"),column.labels = c("HIR", "Higashida"), dep.var.caption = "", title="Predictors of extent of reperfusion",dep.var.labels.include = FALSE, model.numbers = FALSE)
```

### Predictors of clinical outcome
```{r, echo= FALSE, fig.width=4, fig.height=11, fig.align='center',dpi=300, warning=F, results="asis"}
par(mfrow=c(3,2))

model_A <- lmrob(mRS_d90~HIR1, data=allpts, setting = "KS2014")
model_B <- lmrob(mRS_d90~NIHSS_ad + iv_tPR + Rekan*HIR1 + circulation, data=allpts, setting = "KS2014")
model_C <- lmrob(mRS_d90~higashida_D1_bin, data=allpts, setting = "KS2014")
model_D <- lmrob(mRS_d90~NIHSS_ad + iv_tPR + Rekan*higashida_D1_bin + circulation, data=allpts, setting = "KS2014")
model_E <- lmrob(mRS_d90~Tmax_6_reperf_rel, data=allpts, setting = "KS2014")
model_F <- lmrob(mRS_d90~NIHSS_ad + iv_tPR + Rekan*Tmax_6_reperf_rel + circulation, data=allpts, setting = "KS2014")

pdf("mRS_RekanHIR.pdf", width = 5, height = 5)
plot(effect(mod=model_B, term = "Rekan*HIR1"), multiline=T, ci.style="bars")
dev.off()
pdf("mRS_RekanHig.pdf", width = 5, height = 5)
plot(effect(mod=model_D, term = "Rekan*higashida_D1_bin"), multiline=T, ci.style="bars")
dev.off()
pdf("mRS_RekanReperf.pdf", width = 5, height = 5)
plot(effect(mod=model_F, term = "Rekan*Tmax_6_reperf_rel"), multiline=T, ci.style="bars")
dev.off()

stargazer(model_A, model_B, model_C, model_D, model_E, model_F, type="html", align = F, style="all", single.row=T)
stargazer(anova(model_A,model_B), type="html", align = F, style="all", single.row=T, title = "1 vs 2")
stargazer(anova(model_C,model_D), type="html", align = F, style="all", single.row=T, title = "1 vs 3")
stargazer(anova(model_E,model_F), type="html", align = F, style="all", single.row=T, title = "1 vs 4")

stargazer(model_B, model_D, model_F, type="html", align = F, style="all", single.row=T, covariate.labels = c("NIHSS on admission", "IV tPA (yes)", "Recanalization (partial)","Recanalization (full)", "HIR", "Higashida (poor)", "Reperfusion", "Circulation (posterior)", "Recanalization (partial)*HIR", "Recanalization (full)*HIR", "Recanalization (partial)*Higashida (poor)", "Recanalization (full)*Higashida (poor)", "Recanalization (partial)*Reperfusion", "Recanalization (full)*Reperfusion"),column.labels = c("HIR", "Higashida", "Reperfusion"), dep.var.caption = "", title="Predictors of long-term clinical outcome (mRS day 90)",dep.var.labels.include = FALSE, model.numbers = FALSE)
```

### Predictors of absolute infarct growth (D1 - D2)
```{r, echo= FALSE, fig.width=10, fig.height=10, fig.align='center',dpi=300, warning=F, results="asis"}
par(mfrow=c(3,2))

model_A <- lmrob(DWIgrowth~HIR1 + circulation, data=allpts, setting = "KS2014")
model_B <- lmrob(DWIgrowth~NIHSS_ad + iv_tPR + Rekan*HIR1 + circulation, data=allpts, setting = "KS2014")
model_C <- lmrob(DWIgrowth~higashida_D1_bin + circulation, data=allpts, setting = "KS2014")
model_D <- lmrob(DWIgrowth~NIHSS_ad + iv_tPR + Rekan*higashida_D1_bin + circulation, data=allpts, setting = "KS2014")
model_E <- lmrob(DWIgrowth~Rekan*Tmax_6_reperf_rel + circulation, data=allpts, setting = "KS2014")
model_F <- lmrob(DWIgrowth~NIHSS_ad + iv_tPR +Rekan*Tmax_6_reperf_rel + circulation, data=allpts, setting = "KS2014")

pdf("DWIg_RekanHIR.pdf", width = 5, height = 5)
plot(effect(mod=model_B, term = "Rekan*HIR1"), multiline=T, ci.style="bars")
dev.off()
pdf("DWIg_RekanHig.pdf", width = 5, height = 5)
plot(effect(mod=model_D, term = "Rekan*higashida_D1_bin"), multiline=T, ci.style="bars")
dev.off()
pdf("DWIg_RekanReperf.pdf", width = 5, height = 5)
plot(effect(mod=model_F, term = "Rekan*Tmax_6_reperf_rel"), multiline=T, ci.style="bars")
dev.off()

stargazer(model_A, model_B, model_C, model_D, model_E, model_F, type="html", align = F, style="all", single.row=T)
stargazer(anova(model_A,model_B), type="html", align = F, style="all", single.row=T, title = "1 vs 2")
stargazer(anova(model_C,model_D), type="html", align = F, style="all", single.row=T, title = "1 vs 3")
stargazer(anova(model_E,model_F), type="html", align = F, style="all", single.row=T, title = "1 vs 4")

stargazer(model_B, model_D, model_F, type="html", align = F, style="all", single.row=T, covariate.labels = c("NIHSS on admission", "IV tPA (yes)", "Recanalization (partial)","Recanalization (full)", "HIR", "Higashida (poor)", "Reperfusion", "Circulation (posterior)", "Recanalization (partial)*HIR", "Recanalization (full)*HIR", "Recanalization (partial)*Higashida (poor)", "Recanalization (full)*Higashida (poor)", "Recanalization (partial)*Reperfusion", "Recanalization (full)*Reperfusion"),column.labels = c("HIR", "Higashida", "Reperfusion"), dep.var.caption = "", title="Predictors of early infarct growth",dep.var.labels.include = FALSE, model.numbers = FALSE)
```

### Predictors of absolute infarct growth (D1 - D5)
```{r, echo= FALSE, fig.width=10, fig.height=10, fig.align='center',dpi=300, warning=F, results="asis"}
par(mfrow=c(3,2))

model_A <- lmrob(infarct_growth_D5FLAIR~HIR1 + circulation, data=allpts, setting="KS2014")
model_B <- lmrob(infarct_growth_D5FLAIR~NIHSS_ad + iv_tPR + Rekan*HIR1 + circulation, data=allpts, setting="KS2014")
model_C <- lmrob(infarct_growth_D5FLAIR~higashida_D1_bin + circulation, data=allpts, setting="KS2014")
model_D <- lmrob(infarct_growth_D5FLAIR~NIHSS_ad + iv_tPR + Rekan*higashida_D1_bin  + circulation, data=allpts, setting="KS2014")
model_E <- lmrob(infarct_growth_D5FLAIR~Rekan*Tmax_6_reperf_rel + circulation, data=allpts, setting = "KS2014")
model_F <- lmrob(infarct_growth_D5FLAIR~NIHSS_ad + iv_tPR + Rekan*Tmax_6_reperf_rel + circulation, data=allpts, setting = "KS2014")

pdf("FLAIRg_RekanHIR.pdf", width = 5, height = 5)
plot(effect(mod=model_B, term = "Rekan*HIR1"), multiline=T, ci.style="bars")
dev.off()
pdf("FLAIRg_RekanHig.pdf", width = 5, height = 5)
plot(effect(mod=model_D, term = "Rekan*higashida_D1_bin"), multiline=T, ci.style="bars")
dev.off()
pdf("FLAIRg_RekanReperf.pdf", width = 5, height = 5)
plot(effect(mod=model_F, term = "Rekan*Tmax_6_reperf_rel"), multiline=T, ci.style="bars")
dev.off()

stargazer(model_A, model_B, model_C, model_D, model_E, model_F, type="html", align = F, style="all", single.row=T)
stargazer(anova(model_A,model_B), type="html", align = F, style="all", single.row=T, title = "1 vs 2")
stargazer(anova(model_C,model_D), type="html", align = F, style="all", single.row=T, title = "1 vs 3")
stargazer(anova(model_E,model_F), type="html", align = F, style="all", single.row=T, title = "1 vs 4")

stargazer(model_B, model_D, model_F, type="html", align = F, style="all", single.row=T, covariate.labels = c("NIHSS on admission", "IV tPA (yes)", "Recanalization (partial)","Recanalization (full)", "HIR", "Higashida (poor)", "Reperfusion", "Circulation (posterior)", "Recanalization (partial)*HIR", "Recanalization (full)*HIR", "Recanalization (partial)*Higashida (poor)", "Recanalization (full)*Higashida (poor)", "Recanalization (partial)*Reperfusion", "Recanalization (full)*Reperfusion"),column.labels = c("HIR", "Higashida", "Reperfusion"), dep.var.caption = "", title="Predictors of total infarct growth",dep.var.labels.include = FALSE, model.numbers = FALSE)
```

### Predictors of collateral flow defined by Higashida score
```{r, echo= FALSE, fig.width=8, fig.height=5, fig.align='center',dpi=300, warning=F, results="asis"}
model_A <- glmrob(higashida_D1_bin~NIHSS_ad + age + sex + TSI + DWIvol_d1 + circulation, data=allpts, family = "binomial")

stargazer(model_A, type="html", align = F, style="all", single.row=T)
```

### Predictors of collateral flow defined by HIR
```{r, echo= FALSE, fig.width=8, fig.height=5, fig.align='center',dpi=300, warning=F, results="asis"}
model_A <- glmrob(HIR1~NIHSS_ad + age + sex + TSI + DWIvol_d1 + circulation, data=allpts, family = "binomial")


if (effplot == "Y") {
plot(allEffects(model_A), multiline=T, ci.style="bars")

}

stargazer(model_A, type="html", align = F, style="all", single.row=T)
```