---
title: "1000plus collateral flow & perfusion project"
author: "Ahmed Khalil"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: 
    toc: true
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# necessary packages
library(robustbase)
library(knitr)
library(xtable)
library(Hmisc)
library(TOSTER)
library(stargazer)
library(summarytools)
library(magrittr)
library(kableExtra)
library(effects)
library(pwr)
library(papeR)

# DEFINE WHETHER YOU WANT EFFECTS PLOTS FOR MIXED MODELS

effplot = "N"

# load data
allpts <- read.table("all_data_v3.csv", header = T, sep=",")
colnames(allpts)[1] <- "ID"

# load volumetric data
auto_pts <- read.table("ALLVOLS_automated_AK_20180605.csv", header=T, sep=";") # patients automatically delineated
manual_pts <- read.table("ALLVOLS_manual_AK_20180605.csv", header=T, sep=";") # patients manually delineated

# remove # of voxels, keep only volumes (in mm3)
auto_pts <- auto_pts[,c(1,3,5,7,9,11,13)]
manual_pts <- manual_pts[,c(1,3,5,7,9,11,13)]

# convert volumes to mL (cm3)
auto_pts[,-1] <- auto_pts[,-1]*0.001
manual_pts[,-1] <- manual_pts[,-1]*0.001

# combine manual and automated volumes
vols_all_pts <- rbind(auto_pts, manual_pts)

# exclude subject 86 (not in Kersten's main dataset)
vols_all_pts <- vols_all_pts[!vols_all_pts$ID=="86",]

# calculate HIR for days 1 and 2 
vols_all_pts$HIR1 <- vols_all_pts$Tmax_D1_.8_volume/vols_all_pts$Tmax_D1_.2_volume
vols_all_pts$HIR2 <- vols_all_pts$Tmax_D2_.8_volume/vols_all_pts$Tmax_D2_.2_volume

# merge volumes and HIR with the rest of the data
allpts <- merge(allpts, vols_all_pts, by = "ID")

# rename Tmax volume variables to include threshold (>6s)
allpts$Tmax_D1_.6_volume <- allpts$Tmax_vol_d1
allpts$Tmax_D2_.6_volume <- allpts$Tmax_vol_d2

allpts$Tmax_vol_d1 <- c()
allpts$Tmax_vol_d2 <- c()
allpts$Tmax_change <- c()
allpts$relTmax_change <- c()

# create change variables for other thresholds (>4s)
allpts$Tmax_4_change <- allpts$Tmax_D2_.4_volume - allpts$Tmax_D1_.4_volume
allpts$relTmax_4_change <- (allpts$Tmax_4_change/allpts$Tmax_D1_.4_volume)*100
  
# create necessary variables (reperfusion / infarct growth binarized cut-offs)
allpts$reperf[allpts$relTmax_4_change<=-50] <- "Yes"
allpts$reperf[allpts$relTmax_4_change>-50] <- "No"
allpts$reperf <- factor(allpts$reperf)

allpts$infarct_growth[allpts$infarct_growth_D5FLAIR>=7] <- "Yes"
allpts$infarct_growth[allpts$infarct_growth_D5FLAIR<7] <- "No"
allpts$infarct_growth <- factor(allpts$infarct_growth)

# recode recanalization variable
allpts$Rekan[allpts$Rekan==0] <- "None" # no recanalization
allpts$Rekan[allpts$Rekan==1] <- "Full" # full recanalization
allpts$Rekan[allpts$Rekan==2] <- "Partial" # partial recanalization
allpts$Rekan <- factor(allpts$Rekan)
allpts$Rekan <- factor(allpts$Rekan,levels(allpts$Rekan)[c(1,3,2)])

# recode other categorical variables
allpts$sex[allpts$sex==1] <- "Male"
allpts$sex[allpts$sex==2] <- "Female"
allpts$sex <- factor(allpts$sex)

allpts$iv_tPR[allpts$iv_tPR==0] <- "No"
allpts$iv_tPR[allpts$iv_tPR==1] <- "Yes"
allpts$iv_tPR[allpts$iv_tPR==2] <- "Yes" # CHECK WITH KERSTEN IF THIS IS CORRECTLY CODED

allpts$iv_tPR <- factor(allpts$iv_tPR)

allpts$circulation[allpts$circulation==1] <- "Anterior"
allpts$circulation[allpts$circulation==2] <- "Posterior"
allpts$circulation <- factor(allpts$circulation)

allpts$higashida_change <- allpts$higashida_D2-allpts$higashida_D1
allpts$higashida_D1 <- factor(allpts$higashida_D1)
allpts$higashida_D2 <- factor(allpts$higashida_D2)
allpts$higashida_change <- factor(allpts$higashida_change)

# remove certain variables
allpts$no_recan_1 <- c()
allpts$VAR00002 <- c()
allpts$TSI_dich <- c()

# merge with variables defining occlusion location
occ_classfn <- read.table("occlusion_classfn.csv", header = T, sep=";")
allpts <- merge(allpts, occ_classfn, by = "ID")

# correctly code NaNs
allpts[allpts=="NaN"]<-NA

# Defines whether (T) or not (F) to show outliers in boxplots
OL = F
```
### General remarks on methods:
**Reperfusion** is defined as a relative Tmax >4s change of >-50%.  
Surrogates for good collateral flow in this study are:  
  * Presence of reperfusion in non-recanalizers  
  * The hypoperfusion intensity ration (HIR) at baseline (lower is better), defined according to:  
*Bang, O. Y. et al. (2008) ‘Determinants of the distribution and severity of hypoperfusion in patients with ischemic stroke.’, Neurology*  
  * The modified ASITN/SIR Collateral Flow Grading System (adapted for DSC-MRI), according to:  
*Campbell BC V, Christensen S, Tress BM, Churilov L, Desmond PM, Parsons MW, Alan Barber P, Levi CR, Bladin C, Donnan GA, Davis SM (2013) Failure of collateral blood flow is associated with infarct growth in ischemic stroke. J Cereb Blood Flow Metab*

The outcome measures in this study are:  
* Modified Rankin Scale at Day 90  
* Extent of reperfusion - relative Tmax >4s change (%)  
* Absolute infarct growth on DWI between days 1 and 2 (mL)  
* Relative infarct growth on DWI between days 1 and 2 (%)  
* Absolute infarct growth between days 1 (DWI) and 5 (FLAIR)  

All regression analyses are performed using a robust GLM ("High Breakdown And High Efficiency Robust Linear Regression") using the R function "lmrob" from the "robustbase" package.  
Citation of package:  
*Martin Maechler, Peter Rousseeuw, Christophe Croux, Valentin Todorov, Andreas Ruckstuhl, Matias Salibian-Barrera, Tobias Verbeke, Manuel Koller, Eduardo L. T. Conceicao  and Maria Anna di Palma (2018). robustbase: Basic Robust Statistics R package version 0.93-0. URL http://CRAN.R-project.org/package=robustbase *   
Citation of robust GLM:  
*Yohai, V. J. (1987). High breakdown-point and high efficiency robust estimates for regression. The Annals of Statistics, 642-656.*

For visualization (not analysis) purposes, outliers are **not** shown on the boxplots. 

**Power analysis:**
`r pwrtest <- pwr.f2.test(u= 11, f2=0.1, sig.level = 0.05, power=0.8)`

Sample size needed to detect minimum R<sup>2</sup>  of `r pwrtest$f2` at significance level of `r pwrtest$sig.level` and power of `r pwrtest$power` with `r pwrtest$u` predictors (including interactions) = `r pwrtest$v`. Performed with the R package "pwr".  
Citation of package:  
*Stephane Champely (2018). pwr: Basic Functions for Power Analysis. R package version 1.2-2. https://CRAN.R-project.org/package=pwr *
  
  


### Descriptive statistics
```{r, echo= FALSE, fig.width=6, fig.height=6, fig.align='center',dpi=300, warning=F, results="asis"}
# descriptive stats
allpts_summ <- allpts[,c(2:13,15:22,29,30,33,34,35)] # define variables for summary
kable(summarize(allpts_summ, type="numeric"))
kable(summarize(allpts_summ, type="factor", variables=c("reperf", "higashida_D1", "higashida_D2", "Rekan", "circulation", "sex", "iv_tPR")))

```
<style>

table, td, th {
  border: none;
  padding-left: 1em;
  padding-right: 1em;
  min-width: 50%;
  margin-left: auto;
  margin-right: auto;
  margin-top: 1em;
  margin-bottom: 1em;
}

</style>

### Are there differences in time from symptom onset to imaging in different recanalization groups?
```{r, echo= FALSE, fig.width=8, fig.height=5, fig.align='center',dpi=300, warning=F}
# test for differences in TSI according to recanalization status
summary(aov(allpts$TSI~allpts$Rekan))
boxplot(allpts$TSI~allpts$Rekan)
```

### What is the association between reperfusion and recanalization?
```{r, echo= FALSE, fig.width=6, fig.height=6, fig.align='center',dpi=300, warning=F}

# plot recanalization according to reperfusion status
barplot(table(allpts$reperf, allpts$Rekan), legend.text = c("No reperfusion", "Reperfusion"), col=c("#1b9e77","#d95f02"), xlab="Recanalization", ylab="Number of patients", xpd=F,args.legend = list(x = "topleft"), ylim=c(0,80))

# cross-tabulation of recanalization/reperfusion
kable(table(allpts$reperf, allpts$Rekan)) %>%
  kable_styling(full_width=F, position="center") %>%
  add_header_above(c(" "=1, "Recanalization"=3)) %>%
  group_rows("Reperfusion",1,2)

# test for differences between recanalization and reperfusion
print(chisq.test(table(allpts$reperf, allpts$Rekan)))

```
There is a statistically significant difference in the prevalence of reperfusion between the recanalization groups (i.e. reperfusion is associated with recanalization grade).

### Is there a difference in the extent of reperfusion between patients with anterior and posterior circulation strokes?
Models (1) & (2) - in non-recanalizers  
Models (3) & (4) - in recanalizers  
Models (2) & (4) - adjusted for IV tPA  
Models (1) & (3) - not adjusted for IV tPA  

**Note: the second iteration of "iv_tPRYes" in the table below refers to the interaction between circulation + IV tPA**  

```{r, echo= FALSE, fig.width=6, fig.height=6, fig.align='center',dpi=300, warning=F, results="asis"}

# test if circulation has influence on reperfusion in non-recanalizers
model_1 <- lmrob(allpts[allpts$Rekan=="None",]$relTmax_4_change~allpts[allpts$Rekan=="None",]$circulation)
# test if circulation has influence on reperfusion in non-recanalizers ADJUSTING FOR iv TPA
model_2 <- lmrob(allpts[allpts$Rekan=="None",]$relTmax_4_change~allpts[allpts$Rekan=="None",]$circulation*allpts[allpts$Rekan=="None",]$iv_tPR)
# test if circulation has influence on reperfusion in recanalizers (full OR partial)
model_3 <- lmrob(allpts[allpts$Rekan=="Full"|allpts$Rekan=="Partial",]$relTmax_4_change~allpts[allpts$Rekan=="Full"|allpts$Rekan=="Partial",]$circulation)
# test if circulation has influence on reperfusion in recanalizers (full OR partial) ADJUSTING FOR iv TPA
model_4 <- lmrob(allpts[allpts$Rekan=="Full"|allpts$Rekan=="Partial",]$relTmax_4_change~allpts[allpts$Rekan=="Full"|allpts$Rekan=="Partial",]$circulation*allpts[allpts$Rekan=="Full"|allpts$Rekan=="Partial",]$iv_tPR)

stargazer(model_1, model_2, model_3,model_4, type="html", align = F, style="all", single.row=T)

boxplot(relTmax_4_change~circulation, data=allpts, xlab="Circulation", ylab="Extent of reperfusion (%)", outline=OL)

```

### Clinical outcome
```{r, echo= FALSE, fig.width=8, fig.height=5, fig.align='center',dpi=300, warning=F, results="asis"}

model_A <- lmrob(mRS_d90~NIHSS_ad + iv_tPR, data=allpts)
boxplot(allpts$mRS_d90~allpts$iv_tPR, ylab="mRS (day 90)", xlab="IV thrombolysis")
boxplot(allpts$NIHSS_ad~allpts$mRS_d90, xlab="mRS (day 90)", ylab="NIHSS on admission")

model_B <- lmrob(mRS_d90~NIHSS_ad + iv_tPR + Rekan*reperf, data=allpts)
boxplot(mRS_d90~Rekan + reperf, data=allpts, col=c("#1b9e77","#1b9e77","#1b9e77","#d95f02","#d95f02","#d95f02"), ylab="mRS day 90", outline=OL, xaxt="n", xlab="Recanalization")
axis(side=1, at=c(1,2,3,4,5,6), labels=c("None","Partial","Full","None","Partial","Full"))
legend("topright", legend = c("Non-reperfused","Reperfused"), fill=c("#1b9e77","#d95f02"))

model_C <- lmrob(mRS_d90~NIHSS_ad + iv_tPR + HIR1, data=allpts)
boxplot(allpts$HIR1~allpts$mRS_d90, xlab="mRS (day 90)", ylab="HIR")

model_D <- lmrob(mRS_d90~NIHSS_ad + iv_tPR + higashida_D1, data=allpts)
boxplot(allpts$mRS_d90~allpts$higashida_D1, ylab="mRS (day 90)", xlab="Higashida score D1", xlim=c(0.5,5.5), outline=OL)

model_E <- lmrob(mRS_d90~NIHSS_ad + iv_tPR + higashida_D2, data=allpts)
boxplot(allpts$mRS_d90~allpts$higashida_D2, ylab="mRS (day 90)", xlab="Higashida score D2", xlim=c(0.5,5.5), outline=OL)

if (effplot == "Y") {
plot(allEffects(model_A), multiline=T, ci.style="bars")
plot(allEffects(model_B), multiline=T, ci.style="bars")
plot(allEffects(model_C), multiline=T, ci.style="bars")
plot(allEffects(model_D), multiline=T, ci.style="bars")
plot(allEffects(model_E), multiline=T, ci.style="bars")

}

stargazer(model_A, model_B, model_C, model_D, model_E, type="html", align = F, style="all", single.row=T)
```

### Extent of reperfusion
```{r, echo= FALSE, fig.width=8, fig.height=5, fig.align='center',dpi=300, warning=F, results="asis"}

model_A <- lmrob(relTmax_4_change~NIHSS_ad + iv_tPR, data=allpts)
boxplot(allpts$relTmax_4_change~allpts$iv_tPR, ylab="Extent of reperfusion", xlab="IV thrombolysis", outline=OL)
plot(allpts$NIHSS_ad,allpts$relTmax_4_change, ylab="Extent of reperfusion", xlab="NIHSS on admission")

model_B <- lmrob(relTmax_4_change~NIHSS_ad + iv_tPR + Rekan, data=allpts)
boxplot(relTmax_4_change~Rekan, data=allpts, ylab="Extent of reperfusion", outline=OL, xlab="Recanalization")

model_C <- lmrob(relTmax_4_change~NIHSS_ad + iv_tPR + HIR1, data=allpts)
plot(allpts$HIR1,allpts$relTmax_4_change, xlab="HIR", ylab="Extent of reperfusion")

model_D <- lmrob(relTmax_4_change~NIHSS_ad + iv_tPR + higashida_D1, data=allpts)
boxplot(allpts$relTmax_4_change~allpts$higashida_D1, ylab="Extent of reperfusion", xlab="Higashida score D1", xlim=c(0.5,5.5), outline=OL)

model_E <- lmrob(relTmax_4_change~NIHSS_ad + iv_tPR + higashida_D2, data=allpts)
boxplot(allpts$relTmax_4_change~allpts$higashida_D2, ylab="Extent of reperfusion", xlab="Higashida score D2", xlim=c(0.5,5.5), outline=OL)

if (effplot == "Y") {
plot(allEffects(model_A), multiline=T, ci.style="bars")
plot(allEffects(model_B), multiline=T, ci.style="bars")
plot(allEffects(model_C), multiline=T, ci.style="bars")
plot(allEffects(model_D), multiline=T, ci.style="bars")
plot(allEffects(model_E), multiline=T, ci.style="bars")

}

stargazer(model_A, model_B, model_C, model_D, model_E, type="html", align = F, style="all", single.row=T)
```

### Relative infarct growth (D1 - D2)
```{r, echo= FALSE, fig.width=8, fig.height=5, fig.align='center',dpi=300, warning=F, results="asis"}
model_A <- lmrob(rel_DWIgrowth~NIHSS_ad + iv_tPR, data=allpts)
boxplot(allpts$rel_DWIgrowth~allpts$iv_tPR, ylab="Relative infarct growth (D1 - D2)", xlab="IV thrombolysis", outline=OL)
plot(allpts$NIHSS_ad,allpts$rel_DWIgrowth, ylab="Relative infarct growth (D1 - D2)", xlab="NIHSS on admission")

model_B <- lmrob(rel_DWIgrowth~NIHSS_ad + iv_tPR + Rekan*reperf, data=allpts)
boxplot(rel_DWIgrowth~Rekan, data=allpts, ylab="Relative infarct growth (D1 - D2)", outline=OL, xlab="Recanalization")

model_C <- lmrob(rel_DWIgrowth~NIHSS_ad + iv_tPR + HIR1, data=allpts)
plot(allpts$HIR1,allpts$rel_DWIgrowth, xlab="HIR", ylab="Relative infarct growth (D1 - D2)")

model_D <- lmrob(rel_DWIgrowth~NIHSS_ad + iv_tPR + higashida_D1, data=allpts)
boxplot(allpts$rel_DWIgrowth~allpts$higashida_D1, ylab="Relative infarct growth (D1 - D2)", xlab="Higashida score D1", xlim=c(0.5,5.5), outline=OL)

model_E <- lmrob(rel_DWIgrowth~NIHSS_ad + iv_tPR + higashida_D2, data=allpts)
boxplot(allpts$rel_DWIgrowth~allpts$higashida_D2, ylab="Relative infarct growth (D1 - D2)", xlab="Higashida score D2", xlim=c(0.5,5.5), outline=OL)

if (effplot == "Y") {
plot(allEffects(model_A), multiline=T, ci.style="bars")
plot(allEffects(model_B), multiline=T, ci.style="bars")
plot(allEffects(model_C), multiline=T, ci.style="bars")
plot(allEffects(model_D), multiline=T, ci.style="bars")
plot(allEffects(model_E), multiline=T, ci.style="bars")
}

stargazer(model_A, model_B, model_C, model_D, model_E, type="html", align = F, style="all", single.row=T)
```

### Absolute infarct growth (D1 - D2)
```{r, echo= FALSE, fig.width=8, fig.height=5, fig.align='center',dpi=300, warning=F, results="asis"}
model_A <- lmrob(DWIgrowth~NIHSS_ad + iv_tPR, data=allpts)
boxplot(allpts$DWIgrowth~allpts$iv_tPR, ylab="Absolute infarct growth (D1 - D2)", xlab="IV thrombolysis", outline=OL)
plot(allpts$NIHSS_ad,allpts$DWIgrowth, ylab="Absolute infarct growth (D1 - D2)", xlab="NIHSS on admission")

model_B <- lmrob(DWIgrowth~NIHSS_ad + iv_tPR + Rekan*reperf, data=allpts)
boxplot(DWIgrowth~Rekan, data=allpts, ylab="Absolute infarct growth (D1 - D2)", outline=OL, xlab="Recanalization")

model_C <- lmrob(DWIgrowth~NIHSS_ad + iv_tPR + HIR1, data=allpts)
plot(allpts$HIR1,allpts$DWIgrowth, xlab="HIR", ylab="Absolute infarct growth (D1 - D2)")

model_D <- lmrob(DWIgrowth~NIHSS_ad + iv_tPR + higashida_D1, data=allpts)
boxplot(allpts$DWIgrowth~allpts$higashida_D1, ylab="Absolute infarct growth (D1 - D2)", xlab="Higashida score D1", xlim=c(0.5,5.5), outline=OL)

model_E <- lmrob(DWIgrowth~NIHSS_ad + iv_tPR + higashida_D2, data=allpts)
boxplot(allpts$DWIgrowth~allpts$higashida_D2, ylab="Absolute infarct growth (D1 - D2)", xlab="Higashida score D2", xlim=c(0.5,5.5), outline=OL)

if (effplot == "Y") {
plot(allEffects(model_A), multiline=T, ci.style="bars")
plot(allEffects(model_B), multiline=T, ci.style="bars")
plot(allEffects(model_C), multiline=T, ci.style="bars")
plot(allEffects(model_D), multiline=T, ci.style="bars")
plot(allEffects(model_E), multiline=T, ci.style="bars")

}

stargazer(model_A, model_B, model_C, model_D, model_E, type="html", align = F, style="all", single.row=T)
```

### Absolute infarct growth (D1 - D5)
```{r, echo= FALSE, fig.width=8, fig.height=5, fig.align='center',dpi=300, warning=F, results="asis"}
model_A <- lmrob(infarct_growth_D5FLAIR~NIHSS_ad + iv_tPR, data=allpts)
boxplot(allpts$infarct_growth_D5FLAIR~allpts$iv_tPR, ylab="Absolute infarct growth (D1 - D5)", xlab="IV thrombolysis", outline=OL)
plot(allpts$NIHSS_ad,allpts$infarct_growth_D5FLAIR, ylab="Absolute infarct growth (D1 - D5)", xlab="NIHSS on admission")

model_B <- lmrob(infarct_growth_D5FLAIR~NIHSS_ad + iv_tPR + Rekan*reperf, data=allpts)
boxplot(infarct_growth_D5FLAIR~Rekan, data=allpts, ylab="Absolute infarct growth (D1 - D5)", outline=OL, xlab="Recanalization")

model_C <- lmrob(infarct_growth_D5FLAIR~NIHSS_ad + iv_tPR + HIR1, data=allpts)
plot(allpts$HIR1,allpts$infarct_growth_D5FLAIR, xlab="HIR", ylab="Absolute infarct growth (D1 - D5)")

model_D <- lmrob(infarct_growth_D5FLAIR~NIHSS_ad + iv_tPR + higashida_D1, data=allpts)
boxplot(allpts$infarct_growth_D5FLAIR~allpts$higashida_D1, ylab="Absolute infarct growth (D1 - D5)", xlab="Higashida score D1", xlim=c(0.5,5.5), outline=OL)

model_E <- lmrob(infarct_growth_D5FLAIR~NIHSS_ad + iv_tPR + higashida_D2, data=allpts)
boxplot(allpts$infarct_growth_D5FLAIR~allpts$higashida_D2, ylab="Absolute infarct growth (D1 - D5)", xlab="Higashida score D2", xlim=c(0.5,5.5), outline=OL)

if (effplot == "Y") {
plot(allEffects(model_A), multiline=T, ci.style="bars")
plot(allEffects(model_B), multiline=T, ci.style="bars")
plot(allEffects(model_C), multiline=T, ci.style="bars")
plot(allEffects(model_D), multiline=T, ci.style="bars")
plot(allEffects(model_E), multiline=T, ci.style="bars")

}

stargazer(model_A, model_B, model_C, model_D, model_E, type="html", align = F, style="all", single.row=T)
```